---
title: "NHIS"
author: "ML"
date: "8/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document summarizes trend analysis conducted exclusively with R, following recommendations in [National Center for Health Statistics Guidelines for Analysis
of Trends](https://www.cdc.gov/nchs/data/series/sr_02/sr02_179.pdf). This case study examines emergency room use among adults aged 18–64, by insurance status, from the National Health Interview Survey (NHIS).

## Obtaining data

First we download the Person and Sample Adult data files from NCHS for the years 2000-2015. This is simplest using the `lodown` package, which must be installed from GitHub using the `devtools` package.

```{r download, eval=FALSE}
library(devtools)
install_github("ajdamico/lodown")
library(lodown)

# download NHIS catalog
nhis_cat <- get_catalog("nhis",
                        output_dir = file.path(path.expand( "~" ), "NHIS"))

# download person and sample adult files from 2000-2015
# data files downloaded to subdirectory "NHIS" of home directory
lodown("nhis", subset(nhis_cat, year %in% 2000:2015 & 
                        type %in% c("personsx", "samadult")),
       output_dir = file.path(path.expand( "~" ), "NHIS"))

```

The 16 data files must be stacked into a single multiyear table.

```{r packages, eval=FALSE}

library(tidyverse)
library(srvyr)
library(polypoly)
```

For all years except 2004, the PSU, stratum and age variables are in the sample adult file. For 2004, they appear in the person file. We first build a table for 2004 containing the variables needed for analysis. All data files are in .rds format.

```{r 2004, eval=FALSE}
# add explanation of variables
person_04 <- readRDS("2004/personsx.rds") %>% 
  select(srvy_yr, hhx, fmx, ends_with("px"), ends_with("pub"), psu, stratum,
         ends_with("chip"), medicaid, private, notcov, age_p)

samadult_04 <- readRDS("2004/samadult.rds") %>% 
  select(srvy_yr, hhx, fmx, ends_with("px"), starts_with("wtfa"), ahernoy2, 
         starts_with("strat"))

nhis_04 <- inner_join(person_04, samadult_04, 
                      by = c("srvy_yr", "hhx", "fmx", "fpx"))
rm(person_04, samadult_04)
```

For the other years

```{r}
# this function reads the person and sample adult files for the years 2000-2003, 2005-2016, and joins them into an annual data table 
readNHIS <- function(yr) {
  person <- readRDS(here("NHIS", "data", paste0("20", yr), "personsx.rds")) %>% 
    select(srvy_yr, hhx, fmx, ends_with("px"), ends_with("pub"), 
           ends_with("chip"), medicaid, private, notcov)
  
  samadult <- readRDS(here("NHIS", "data", paste0("20", yr), 
                           "samadult.rds")) %>% 
    select(srvy_yr, hhx, fmx, ends_with("px"), starts_with("wtfa"), ahernoy2,
           starts_with("psu"), starts_with("strat"), age_p,
           ends_with("chip"), ends_with("pub"))
  
  inner_join(person, samadult)
  }

nhis_dat <- map_dfr(sprintf("%02d", (0:15)[-5]), readNHIS) %>% 
  unnest(cols = srvy_yr) %>% 
  # add 2004 to table and arrange in year order
  bind_rows(nhis_04) %>% 
  arrange(srvy_yr, hhx)
```

